---
title: "Change in US Renewable Energy Share (2021-2023) - Lab 4 Project Report"
author: "Emma Mott"
format: typst
---


## **Part 0: libraries**

```{r, echo = 'true', results= 'hide'}

#install.packages("sf")
library("sf")
library("tidyverse")
#install.packages("rnaturalearth")
library("rnaturalearth")
library(stringr)
library(dplyr)
library(ggplot2)
#install.packages("maps")
library(maps)

```

## **Part 1:** **Defining Research Question**

## Overview

Chosen Question: How does the share of renewable energy use across states change between 2021-2023?

## Data and Methods

## **Part 2: Data Preparation and Cleaning**

```{r}

#Load data for renewables
renew_2021 <- read_csv("data/renew-use-2021.csv")
renew_2022 <- read_csv("data/renew-use-2022.csv")
renew_2023 <- read_csv("data/renew-use-2023.csv")

# renaming columns
renew_2021 <- renew_2021 |> rename(Renewable_Use = matches("Renewable_Use"))
renew_2022 <- renew_2022 |> rename(Renewable_Use = matches("Renewable_Use"))
renew_2023 <- renew_2023 |> rename(Renewable_Use = matches("Renewable_Use"))

# adding year column
renew_2021 <- renew_2021 |> mutate(Year = 2021)
renew_2022 <- renew_2022 |> mutate(Year = 2022)
renew_2023 <- renew_2023 |> mutate(Year = 2023)

# combining all three years into one data frame
renew_all <- bind_rows(renew_2021, renew_2022, renew_2023)

# extracting only the numeric values from messy Renewable_Use column using stringr
renew_all <- renew_all |>
  mutate(Renewable_Use_Num = as.numeric(str_extract(Renewable_Use, "\\d+")))

# summarize total renewable use by state and year
renew_summary <- renew_all |>
  group_by(State, Year) |>
  summarise(Total_Renewable_Use = sum(Renewable_Use_Num, na.rm = TRUE)) |>
  mutate(State = str_to_upper(State))

# Load total energy use
total_2021 <- read_csv("data/total-use-2021.csv")
total_2022 <- read_csv("data/total-use-2022.csv")
total_2023 <- read_csv("data/total-use-2023.csv")

head(renew_summary)


```



## **Part 3: Joining / Pivoting Datasets for Analysis**

```{r}

# Pivot longer for 2021–2023
pivot_total <- function(df, year) {
  df |>
    pivot_longer(
      cols = -Energy_Source,
      names_to = "State",
      values_to = "Energy_Use"
    ) |>
    mutate(Year = year)
}

total_2021_long <- pivot_total(total_2021, 2021)
total_2022_long <- pivot_total(total_2022, 2022)
total_2023_long <- pivot_total(total_2023, 2023)

# Combine all years
total_all <- bind_rows(total_2021_long, total_2022_long, total_2023_long)

# Make all state names uppercase
total_all <- total_all |> mutate(State = str_to_upper(State))

# Compute total energy per state 
total_all_summary <- total_all |>
  group_by(State, Year) |>
  summarise(Total_Energy_Use = sum(Energy_Use, na.rm = TRUE))

# Compute total renewable energy per state
renewable_all <- total_all |>
  filter(Energy_Source %in% c(
    "total_renewable_energy", 
    "total_renewables", 
    "total renewable-energy")) |>
  select(State, Year, Renewable_Energy_Use = Energy_Use)

# join them to compute renewable share properly
energy_joined <- total_all_summary |>
  left_join(renewable_all, by = c("State", "Year")) |>
  mutate(Renewable_Share = Renewable_Energy_Use / Total_Energy_Use)

# Convert state abbreviations to full names
state_fulls <- c(state.name, "District of Columbia")
state_abbs  <- c(state.abb, "DC")
# Build lookup table
state_lookup <- tibble(
  state_abbr = state_abbs,   
  state_name = state_fulls     
)


#  clean abbreviated states
energy_clean <- energy_joined |>
  left_join(state_lookup, by = c("State" = "state_abbr")) |>
  mutate(State = state_name) |>
  select(-state_name)

#final clean table with all the data!!
head(energy_clean)


```

## **Part 4: Mapping Visualization**

```{r}


# loading US states geometry and excluding territories
us_states <- ne_states(country = "United States of America", returnclass = "sf") |>
  filter(!name %in% c("Puerto Rico", "Guam", "American Samoa",
                      "Northern Mariana Islands", "United States Virgin Islands"))

# Join the data to the sf object by full state name
us_map <- us_states |>
  left_join(energy_clean, by = c("name" = "State"))

#plotting renewable share over the years
ggplot(us_map) +
  geom_sf(aes(fill = Renewable_Share), color = "white", size = 0.2) +
  facet_wrap(~Year) +
  scale_fill_viridis_c(
  name = "Share of total energy",
  labels = scales::percent_format(accuracy = 1),
  option = "C",
  limits = c(0, 0.4)  # fix the same scale across facets
)+
  coord_sf(
    xlim = c(-125, -66),   # restricts to continental U.S.
    ylim = c(24, 50),      # zooms in to main latitudes
    expand = FALSE
  ) +
  theme_minimal() +
  labs(
    title = "Renewable Energy Share by State",
  ) +
  theme(
    panel.grid.major = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "right"
  )



```


## Analysis 
From the map, we can see that most states do not have a significant change in their renewable energy share of total energy between 2021-2023. However, we do see that there are signifcant differences in renewable energy share across each state. 

It seems that the West coast and midwest states have higher renewable shares than the South or Northeast (beside Maine and Vermont, which also have high renewable shares.) Significant West coast states with large shares are Oregon, Washington, and Idaho. Significant midwest states are South Dakota, Nebraska, and Iowa. In fact, South Dakota consistently has the largest renewable energy share of all the states, indicated by the bright yellow color. 

Another interesting trend is that California's renewable energy share did grow from 2021-2023, indicated by the lighter purple shade in 2023. It is also interesting that the California renewable energy share is comparatively lower than other states, as California is known to produce a large amount of renewable energy. 